---
title: "Getting started with nimbleAPT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Add this vignette to the documentation via roxygen2::roxygenise() and  R CMD build -->

This vignette assumes the reader is already familiar with using nimble to perform MCMC.

The principal motivation of this package is that, when target posterior distributions are multimodal standard Markov chain Monte Carlo (MCMC) algorithms often perform badly.
In these situations, standard algorithms often propose few, if any, jumps between modes, and can thus provide very poor approximation to the target posterior distribution.
The nimbleAPT package permits nimble users to perform adaptive parallel tempering (APT) as a potential solution for sampling multimodal posterior distributions.

Parallel tempering is an MCMC technique where the posterior likelihood of a Bayesian model is 'heated up' to various degrees.
As temperature is increased, a likelihood will become flatter and this can greatly increase the probability and frequency of between-mode jumps. 
In practice, a "temperature ladder" is established, MCMC is performed within each rung of the temperature ladder, and a special between-rung MCMC step is introduced so that parameter sets can move up and down the temperature ladder.
Bayesian inference remains valid so long as it is made on the posterior samples of the unheated rung only.
Adaptive parallel tempering algorithms automaticly tune the tempertures of the temperature ladder so that target acceptance rates for proposed between-rung jumps are achieved. 

The nimbleAPT package provides the following...
1. A buildAPT function: adapted from nimble's buildMCMC, this function sets up the APT algorithm, including between-rung steps and to adapt the temperature ladder adaptation.
2. A set of samplers adapted from nimble's MCMC to include heating of the likelihood.






```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nimbleAPT)
```

```{r}
2 + 2
```
